---
title: "DS 2003 Midterm Report"
subtitle: "Exploring NBA 2021-2022 Stats"
author: "Kevin Ma, Eva Massarelli, Hank Dickerson"
date: "3-17-2023"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cosmo
    code_folding: hide
---
# Executive Summary
This report utilizes NBA data to create clear and informative plots that demonstrate differences in NBA seasons and positions. This report is geared towards readers looking to understand more about how different players perform in the playoffs vs during the regular season and which positions in the NBA let players have the most playtime. Stakeholders interested in this may be recruiters, coaches, or fantasy league players.

Our key findings were that though, average performance between player's performance in the playoffs and regular season were close, players tended to play slightly worse in the regular season than in the playoffs. There are many reasons for this which could include the playoffs being very high stakes as the NBA title is determined by them and that there are fewer games than in the regular season. We also found that the positions of power-forward and centers played the most games but also had the least playing time per game. We found this interesting and thought it may be a valuable thing for coaches to know as they plan substitutes.


# Report Dependencies
```{r, class.source = 'fold-show', message=FALSE}
library(tidyverse)
library(ggplot2)
library(reshape2)
library(skimr)
library(pollster)
library(kableExtra)
```


# Data Set
Our data, ["2021-2022 NBA Player Stats,"](https://www.kaggle.com/datasets/vivovinco/nba-player-stats) was found on Kaggle and is owned by Vivo Vinco. We accessed the data through mid-February to mid-March.


### Raw Data Set
The data set provided by Kaggle were split into data sets. One for the regular and the playoffs season.

<br>
<br>

#### Playoffs Season Data Set
```{r}
# Import Data Set Playoffs Season
playoffs <- read.csv('2021-2022 NBA Player Stats - Playoffs.csv')
playoffs %>%
  kbl() %>%
  kable_styling("hover") %>%
  scroll_box(width = "800px", height = "200px")
```

<br>
<br>

#### Regular Season Data Set
```{r}
# Import Data Set Regular Season
regular <- read.csv('2021-2022 NBA Player Stats - Regular.csv')
regular %>%
  kbl() %>%
  kable_styling("hover") %>%
  scroll_box(width = "800px", height = "200px")
```

<br>
<br>

#### Pertinent Data Dictionary
|Abbreviation|Definition|
|------------|-------|
|Rk|Rank|
|Player|Player's Name|
|Pos|Position|
|Age|Player's age|
|Tm|Team|
|G|Games played|
|GS|Games started|
|MP|Minutes played per game|
|FT|Free throws per game|
|TRB|Total rebounds per game|
|ASR|Assists per game|
|STL|Steals per game|
|BLK|Blocks per game|
|TOV|Turnovers per game|
|PF|Personal fouls per game|
|PTS|Points per game|

<br>
<br>

### Data transformations used for Facet-Wrapped Scatterplot's Data Dictionary

Manipulations performed:

* Combined the playoffs and regular season data with .playoffs or .regular as distinctions
* Filtered to include 15th highest average in either the playoffs or regular season
* Fixed names of players who have accents
* Manipulated data dictionary to be in tidy format to be used for ggplot
* Pivot the data to a wide format that can be used with geom_point()
* Set levels in the desired order
```{r warning=FALSE}
# Combine playoffs and regular season with .playoffs or .regular as distinctions
nba <- playoffs %>%
  left_join(regular, by="Player", suffix = c(".playoffs", ".regular"))
  
nba <- playoffs %>%
  left_join(regular, by="Player", suffix = c(".playoffs", ".regular"))
# filter to only include nba players who average more than:
# top15th leader in ppg in the regular or playoff or
# top15th leader in rpg in the regular or playoff or
# top15th leader in apg in the regular or playoff
nba_filtered <- nba %>%
  group_by(Player) %>%
  summarize(
    PPG_regular = round(mean(PTS.regular, na.rm = TRUE), 2),
    RPG_regular = round(mean(TRB.regular, na.rm = TRUE), 2),
    APG_regular = round(mean(AST.regular, na.rm = TRUE), 2),
    PPG_playoffs = round(mean(PTS.playoffs, na.rm = TRUE), 2),
    RPG_playoffs = round(mean(TRB.playoffs, na.rm = TRUE), 2),
    APG_playoffs = round(mean(AST.playoffs, na.rm = TRUE), 2)
  ) %>%
  filter(
    PPG_regular >= 24.6 | PPG_playoffs >= 22.8 |
    RPG_regular >= 10.1 | RPG_playoffs >= 7.6 |
    APG_regular >= 7 | APG_playoffs >= 5.9
  )
# Manually fix players who have accents in their name
nba_filtered$Player[22] <- "Jonas Valanciunas"
nba_filtered$Player[30] <- "Luka Doncic"
nba_filtered$Player[32] <- "Nikola Jokic"
nba_filtered$Player[33] <- "Nikola Vucevic"

# tidy format that can be plotted with ggplot
nba_tidy <- nba_filtered %>%
  select(Player, PPG_regular, RPG_regular, APG_regular, PPG_playoffs, RPG_playoffs, APG_playoffs) %>%
  gather(key, value, -Player) %>%
  separate(key, into = c("Attribute", "Season"), sep = "_") %>%
  spread(Season, value)

# pivot the data to a wide format that can be used with geom_point()
nba_pivot<- nba_tidy %>%
  pivot_longer(cols = c("regular", "playoffs"), names_to = "Season", values_to = "amount") %>%
  separate(Attribute, into = c("Stat", "Type"), sep = "_") %>%
  pivot_wider(names_from = Season, values_from = amount)
# set the levels in desired order
nba_pivot$Stat <- factor(nba_pivot$Stat, levels = c("PPG", "RPG", "APG"))
```

### Data transformations used for Box plot's Data Dictionary

Manipulations Performed:

* Used regular season dataset as it has more games and includes the whole NBA
* Selected Pos and G for boxplot1
* Selected Pos and MP for boxplot2
* Renamed Pos to Position for both boxplots
* Renamed G to Games for boxplot1
* Renamed MP to Minutes Played for boxplot2
```{r}
# Box plot 1 shows the amount of games each position plays
regular_b1<-regular %>%
  select(Pos, G) %>% # we only need these two columns
  rename(Position = Pos, Games_Played = G)

# Box plot 2 shows the minutes played per game by each position
regular_b2<-regular %>%
  select(Pos, MP) %>% # we only need these two columns
  rename(Position = Pos, Minutes_Played_Per_Game = MP)
```


# Background/Questions
As a group we found that we all had a common interest in basketball, especially with March Madness approaching. Additionally, it is simple to find stats tracking performance of players in sports. Knowing this, we explored some datasets on Kaggle and found an interesting one that included NBA stats on playoffs and regular seasons from 2021-2022.

We thought it would be interesting to see how a player's performance differs between the playoffs and regular seasons. The regular season is obviously a major part of the season and getting into the playoffs but we felt it was important to see how players stepped up to the pressure of being in the playoffs and winning when it mattered.This led to us wondering if there was a difference between a player’s position and how long they are able to remain in the league. For example, players who play a guard position are oftentimes shorter than someone who plays a position like a center. In general those who are taller seem to be less durable and we wanted to see if this applied in the NBA.


# Exploratory Data Anaylsis

#### Combined Data Set
```{r}
# Combining playoffs and regular season with .playoffs or .regular as distinctions
nba <- playoffs %>%
  left_join(regular, by="Player", suffix = c(".playoffs", ".regular"))
nba %>%
  kbl() %>%
  kable_styling("hover") %>%
  scroll_box(width = "800px", height = "200px")
```

<br>
<br>

#### Player's PPG, RPG, and APG in Regular v.s. Playoffs Season
```{r}
nba_filtered %>%
  kbl() %>%
  kable_styling("hover") %>%
  scroll_box(width = "800px", height = "200px")
```

<br>
<br>

#### To make summary statistics
Made df with combined Playoffs and Regular season PTS, AST, and TRB data for players who were in both.
```{r}
# first join dfs to delete players in the regular season that do not appear in the playoffs
joined_filtered <- merge(playoffs, regular, by='Player')

# now we made a subset of the joined dataframes that contains the relevant categories we want to look at for "success" in the NBA.
joined_subset <- joined_filtered %>%
  select(Player, Pos.x, MP.x, MP.y,
         PTS.x, AST.x, TRB.x,
         PTS.y, AST.y, TRB.y,) %>%
  rename(
    Position = Pos.x,
    Playoff_Mins_Played = MP.x,
    Regular_Mins_Played = MP.y,
    Playoffs_PTS = PTS.x, 
    Playoffs_AST = AST.x, 
    Playoffs_TRB = TRB.x, 
    Regular_PTS = PTS.y, 
    Regular_AST = AST.y, 
    Regular_TRB = TRB.y) %>%
  distinct(Player, .keep_all = TRUE)
joined_subset$Position<- as.factor(joined_subset$Position)

# Normalize statistics to minutes played
joined_norm <- joined_subset %>%
  mutate(
    Playoffs_PTS = Playoffs_PTS/Playoff_Mins_Played, 
    Playoffs_AST = Playoffs_AST/Playoff_Mins_Played, 
    Playoffs_TRB = Playoffs_TRB/Playoff_Mins_Played, 
    Regular_PTS = Regular_PTS/Regular_Mins_Played, 
    Regular_AST = Regular_AST/Regular_Mins_Played, 
    Regular_TRB = Regular_TRB/Regular_Mins_Played) %>%
  select(Player, Position,
         Playoffs_PTS, Regular_PTS,
         Playoffs_AST, Regular_AST,
         Playoffs_TRB, Regular_TRB)

# Take percent difference between playoff and regular performance for each player
Percent_Diff <- joined_norm %>%
  group_by(Player) %>%
  summarize(
    "Percent Difference Points (%)" = ((abs(Playoffs_PTS-Regular_PTS))/((Playoffs_PTS+Regular_PTS)/2))*100,
    "Percent Difference Assists (%)" = ((abs(Playoffs_AST-Regular_AST))/((Playoffs_AST+Regular_AST)/2))*100,
    "Percent Difference Rebounds (%)" = ((abs(Playoffs_TRB-Regular_TRB))/((Playoffs_TRB+Regular_TRB)/2))*100) %>%
  mutate_if(is.double, round, digits = 2) 
Percent_Diff$Position <- joined_subset$Position

Percent_Diff %>%
  kbl() %>%
  kable_styling("hover") %>%
  scroll_box(width = "800px", height = "200px")
```

### Summary Statistics and Data Descriptives for Question 1 {.tabset}
#### Question 1 Data Descriptives
```{r}
summary(joined_norm)
```
#### Percent Difference in Playoffs vs Regular Season by Player
```{r}
Percent_Diff_by_Players <- Percent_Diff %>%
  select(Player, `Percent Difference Points (%)`,`Percent Difference Assists (%)`, `Percent Difference Rebounds (%)`)
Percent_Diff_by_Players %>%
  kbl() %>%
  kable_styling("hover") %>%
  scroll_box(width = "800px", height = "200px")
```
#### Percent Difference in Playoffs vs Regular Season by Position
```{r}
Percent_Diff_by_Position <- Percent_Diff %>%
  group_by(Position) %>%
  summarize(
    "Mean Percent Difference Points (%)" = mean(`Percent Difference Points (%)`),
    "Mean Percent Difference Assists (%)" = mean(`Percent Difference Assists (%)`),
    "Mean Percent Difference Rebounds (%)" = mean(`Percent Difference Rebounds (%)`)) %>%
  mutate_if(is.double, round, digits = 2)
  
Percent_Diff_by_Position %>%
  kbl() %>%
  kable_styling("hover") %>%
  scroll_box(width = "800px", height = "200px")
```
### {-}


Question 2 uses raw data as it only compares two variables at a time. Because of this, tidying was not necessary.

### Summary Statistics and Data Descriptives for Question 2 {.tabset}
#### Data Descriptives for Question 2
```{r}
regular_subset <- regular %>%
  select(Player, Pos, G, MP) %>%
  rename(Position = Pos)
summary(regular_subset)
```
#### Data Descriptives for all variables of regular season
```{r}
summary(regular)
```
#### Summary Statistics
```{r}
regular_subset <- regular %>%
  select(Player, Pos, G, MP) %>%
  rename(Position = Pos) %>%
  group_by(Position) %>%
  summarize("Average Time Played Per Game" = mean(MP),
         "Average Games Played" = mean(G))

regular_subset %>%
  kbl() %>%
  kable_styling("hover") %>%
  scroll_box(width = "800px", height = "200px")
```
### {-}

# Question 1 Plot & Narrative
### Question
> How does NBA success translate from playoffs to the regular season in terms of PPG, RPG, AND APG?

### Plot

<br>
<br>

```{r}
# plot facet-wrapped graph
ggplot(nba_pivot, aes(Player, regular, color = "Regular")) +
  geom_point() +
  geom_point(aes(Player, playoffs, color = "Playoff")) +
  facet_grid(Stat ~ ., scales = "free_y") +
  scale_color_manual(name = "Season", values = c("Regular" = "steelblue", "Playoff" = "red")) +
  labs(title = "Regular Season vs Playoff Season Averages") +
  labs(x = "", y ="") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```          

### Narrative
Despite the few outliers most players performed close to their regular season averages in the playoffs. Although most of the averages went down from the regular season to the playoffs. This could contribute to the fact that the level of play increases in the playoffs, especially defensively, as the playoffs are what truly matters when it comes to winning. If the NBA wants all players to play at their peak level throughout the season and not just the playoffs then they might want to think about reducing the amount of games during the regular season. This graph is mainly directed towards NBA analysts and NBA coaches. A lot of analysts discuss who will show up in the playoffs and if they will fail to perform under the pressure. These graphs could provide good insight on who might over perform or under perform when it comes to the playoffs. This graph may also provide good information to coaches when considering who to trade for or sign later in the season if they want a better chance at winning the NBA finals. 

<br>
<br>

# Question 2 Plot & Narrative
### Question
> What positions are the most durable (played for the longest and for the most games) in a regular season?

### Plot

<br>
<br>

#### Box plot of games played for each position (Regular season)
```{r}
#boxplot
ggplot(regular_b1, 
       aes( x = Position, 
            y = Games_Played,
            )) +
  geom_boxplot(color='#0B1D51', fill='steelblue', alpha=0.5) +
    labs(
    title = 'Games Played Based on Position',
    caption = 'Kaggle, 2021-2022 NBA Player Stats
    Center = C
    Center-Power Forward = C-PF
    Power Forward = PF
    Power Forward-Small Forward = PF-SF
    Point Guard = PG
    Point Guard-Shooting Guard = PG-SG
    Shooting Guard = SG
    Shooting Guard-Point Guard = SG-PG
    Shooting Guard-Small Forward = SG-SF') +
  xlab('Position') +
  ylab('Games Played') +
  theme_minimal()
```

<br>
<br>

#### Box plot of minutes played for each position (Regular season)
```{r}
#boxplot
ggplot(regular_b2, 
       aes( x = Position, 
            y = Minutes_Played_Per_Game,
            )) +
  geom_boxplot(color='#0B1D51', fill='steelblue', alpha=0.5) +
    labs(
    title = 'Minutes Played Per Game Based on Position',
    caption = 'Kaggle, 2021-2022 NBA Player Stats
    Center = C
    Center-Power Forward = C-PF
    Power Forward = PF
    Power Forward-Small Forward = PF-SF
    Point Guard = PG
    Point Guard-Shooting Guard = PG-SG
    Shooting Guard = SG
    Shooting Guard-Point Guard = SG-PG
    Shooting Guard-Small Forward = SG-SF') +
  xlab('Position') +
  ylab('Minutes Played per Game') +
  theme_minimal()
```


### Narrative
We thought that it would be good to see what positions played the most games per season to test durability. While this was a good graph we also were curious about the minutes played per game, due to the fact that someone may play a lot of games despite playing few minutes per game. These two box plots did present interesting data as Shooting Guard-Point Guard position was a position that played a lot of games per season on average compared to many others, but despite that the median amount of minutes played per game was much closer to other positions medians. This data could have big implications on the stakeholders (the NBA coaches). These graphs can tell the coaches that although some positions could be the most important part of a team, a team may also want the most substitutions for those respective positions as they are the least durable/can handle less play time throughout a single game.

# Summary
**insert summary paragraphs**

# Citations

**Source for mutate_if function used in summary statistics:**
Team, D. S. T. (2021, September 13). How to Round Data Frame Containing Character Variables in R. Universe of Data Science. Retrieved March 16, 2023, from https://universeofdatascience.com/how-to-round-data-frame-containing-character-variables-in-r/#more-913 

**Source for ending tabset:**
Christophe Dervieux, Yihui Xie, & Emily Riederer. (2022, November 7). R Markdown Cookbook. 7.6 Put content in tabs. Retrieved March 16, 2023, from https://bookdown.org/yihui/rmarkdown-cookbook/html-tabs.html 
