---
title: "Summary Statistics"
output:
  html_document:
    df_print: paged
---

Import dependencies 
```{r}
library(tidyverse)
library(skimr)
library(pollster)
library(kableExtra)
```

Import dataset
```{r}
playoffs <- read.csv('2021-2022 NBA Player Stats - Playoffs.csv')
regular <- read.csv('2021-2022 NBA Player Stats - Regular.csv')
```

Check dataset
```{r}
head(playoffs)
head(regular)
```

Make df with combined Playoffs and Regular season PTS, AST, TRB, and STL data for players who were in BOTH
```{r}
# first join dfs to delete players in the regular season that do not appear in the playoffs
joined_filtered <- merge(playoffs, regular, by='Player')
joined_filtered
joined_subset <- joined_filtered %>%
  select(Player, Pos.x, MP.x, MP.y,
         PTS.x, AST.x, TRB.x, STL.x,
         PTS.y, AST.y, TRB.y, STL.y) %>%
  rename(
    Position = Pos.x,
    Playoff_Mins_Played = MP.x,
    Regular_Mins_Played = MP.y,
    Playoffs_PTS = PTS.x, 
    Playoffs_AST = AST.x, 
    Playoffs_TRB = TRB.x, 
    Playoffs_STL = STL.x,
    Regular_PTS = PTS.y, 
    Regular_AST = AST.y, 
    Regular_TRB = TRB.y, 
    Regular_STL = STL.y) %>%
  distinct(Player, .keep_all = TRUE)

joined_subset$Position<- as.factor(joined_subset$Position)

joined_subset
```

Normalize statistics to minutes played
```{r}
joined_norm <- joined_subset %>%
  mutate(
    Playoffs_PTS = Playoffs_PTS/Playoff_Mins_Played, 
    Playoffs_AST = Playoffs_AST/Playoff_Mins_Played, 
    Playoffs_TRB = Playoffs_TRB/Playoff_Mins_Played, 
    Playoffs_STL = Playoffs_STL/Playoff_Mins_Played,
    Regular_PTS = Regular_PTS/Regular_Mins_Played, 
    Regular_AST = Regular_AST/Regular_Mins_Played, 
    Regular_TRB = Regular_TRB/Regular_Mins_Played, 
    Regular_STL = Regular_STL/Regular_Mins_Played) %>%
  select(Player, Position,
         Playoffs_PTS, Regular_PTS,
         Playoffs_AST, Regular_AST,
         Playoffs_TRB, Regular_TRB,
         Playoffs_STL, Regular_STL)
joined_norm
```

Take percent difference between playoff and regular performance for each player
```{r}
Percent_Diff <- joined_norm %>%
  group_by(Player) %>%
  summarize(
    "Percent Difference Points (%)" = ((abs(Playoffs_PTS-Regular_PTS))/((Playoffs_PTS+Regular_PTS)/2))*100,
    "Percent Difference Assists (%)" = ((abs(Playoffs_AST-Regular_AST))/((Playoffs_AST+Regular_AST)/2))*100,
    "Percent Difference Rebounds (%)" = ((abs(Playoffs_TRB-Regular_TRB))/((Playoffs_TRB+Regular_TRB)/2))*100,
    "Percent Difference Steals (%)" = ((abs(Playoffs_STL-Regular_STL))/((Playoffs_STL+Regular_STL)/2))*100) %>%
  mutate_if(is.double, round, digits = 2) #source: https://universeofdatascience.com/how-to-round-data-frame-containing-character-variables-in-r/#more-913
Percent_Diff$Position <- joined_subset$Position
Percent_Diff
```


Generate summaries
```{r}
summary(joined_subset)
summary(Percent_Diff)
```

## Summarize the pertinent data using descriptive statistics.
Your analysis should include a robust summary of the pertinent variables from your 
dataset. Use well formatted tables and data viz to show your summary descriptives

## Summary Statistics {.tabset}

### 1) How does NBA success translate from playoffs to the regular season? {.tabset}

#### Percent Difference by Player
```{r}

Percent_Diff_by_Players <- Percent_Diff %>%
  select(Player, `Percent Difference Points (%)`,`Percent Difference Assists (%)`, `Percent Difference Rebounds (%)`, `Percent Difference Steals (%)`)

Percent_Diff_by_Players %>%
  kbl() %>%
  kable_styling("hover") %>%
  scroll_box(width = "800px", height = "200px")
```

#### Percent Difference by Position {.tabset}
```{r}
Percent_Diff_by_Position <- Percent_Diff %>%
  group_by(Position) %>%
  summarize(
    "Mean Percent Difference Points (%)" = mean(`Percent Difference Points (%)`),
    "Mean Percent Difference Assists (%)" = mean(`Percent Difference Assists (%)`),
    "Mean Percent Difference Rebounds (%)" = mean(`Percent Difference Rebounds (%)`),
    "Mean Percent Difference Steals (%)" = mean(`Percent Difference Steals (%)`)) %>%
  mutate_if(is.double, round, digits = 2)
  
Percent_Diff_by_Position %>%
  kbl() %>%
  kable_styling("hover") %>%
  scroll_box(width = "800px", height = "200px")
```


### 2) What positions are the most durable in a regular season? {.tabset}
```{r}
regular_subset <- regular %>%
  select(Player, Pos, G, MP) %>%
  group_by(Pos) %>%
  summarize("Average Time Played Per Game" = mean(MP),
         "Average Games Played" = mean(G))
regular_subset %>%
  kbl() %>%
  kable_styling("hover") %>%
  scroll_box(width = "800px", height = "200px")
```

-Box plot of games played for each position (Regular season)
-Box plot of minutes played for each position (Regular season)
