---
title: "DS2003 Team Group Project"
author: "Kevin, Hank, Eva"
date: "2023-02-20"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cosmo
    code_folding: show
---

# Import Dependencies
```{r message = FALSE}
library(tidyverse)
library(ggplot2)
```

# Questions
1) Do players perform better in the playoffs than the regular seasons in terms of PPG?
2) Who is the most durable player in terms of MPG and GP?

# Import Data
```{r}
regular <- read.csv("/Users/kevinma/Desktop/UVA/DS2003/Team Group Project/2021-2022 NBA Player Stats - Regular.csv")
playoffs <- read.csv("/Users/kevinma/Desktop/UVA/DS2003/Team Group Project/2021-2022 NBA Player Stats - Playoffs.csv")
```

# Pull and Clean Dataset
```{r}
# Combining playoffs and regular season with .playoffs or .regular as distinctions
nba <- playoffs %>%
  left_join(regular, by="Player", suffix = c(".playoffs", ".regular"))

head(nba, 6)
```

# Which players perform better or worse in the playoffs than the regular seasons in terms of PPG?
```{r}
# Select Players who average >=20 PPG in the playsoffs and regular season to cut down the entries
points <- nba %>% 
  select(Player, PTS.regular, PTS.playoffs) %>%
  filter(PTS.regular >= 20, PTS.playoffs >=20) %>%
  group_by(Player) %>%
  # If a player has multiple entries due to being traded, combine through finding the mean for the PPG columns
  summarize(PTS.regular = mean(PTS.regular), PTS.playoffs = mean(PTS.playoffs))
```

```{r}
ggplot(points, aes(x = Player, y= PTS.playoffs)) + 
  # Red = playoffs
  geom_point(color = "red") + 
  # Blue = regular season
  geom_point(aes(y = PTS.regular), color = "blue") + 
  # Add labels
  labs(title = "Player and PPG in the Regular Season (Blue) and Playsoff Season (Red)",
       x = "Player", y = "PPG >=20") + theme_minimal() + 
  # Rotate Player Names by 90 degrees
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

- Hank: Cleaned the dataset and added comments 
- Kevin: Created the scatterplot and made the scatterplot readable
- Ana: Cleaned the dataset and helped filter and sort the necessary columns for the scatterplot

